---
title: "final_mbb_alexithymia_analyses"
author: "FQ"
date: "2026-04-21"
output: html_document
---
Final sharing version of script for paper: <title>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
## load libraries
```{r}
suppressPackageStartupMessages({
  library(sjPlot)
  library(ggplot2)
  library(tidyverse)
  library(car) #for ANCOVA
  library(stats) #for ANOVA
  library(multcomp) #for ANCOVA pairwise comparisons
  library(ggsignif) #for significance stars on graphs
  library(ggformula) #for regression diagnostics
  library(ggExtra) #for marginal distributions on scatterplot
  library(Hmisc) # for many histograms at once
  library(lavaan) # for SEM LGM modeling
  library(multiplex) # for writing .dat files
  library(patchwork) # for combining plots
  library(haven) # for reading in SAV files
  library(lavaanPlot)
  library(corrplot) # for correlation matrices
  library(lme4)
  library(gtsummary) # for descriptive stats
  library(gt) # for saving tables
  library(rstatix)
})
```

## load data
```{r}
data <- read_csv("ela_ss_alexithymia_data.csv")

# updated alexithymia data pulled in wrangling script already has this
#alexi_w1 <- read_csv("../../Wave_1/Data_scoring/Alexithymia/alexithymia_scores_w1_20231017.csv") %>% dplyr::select(participant_id, contains("idcom"))

# alexithymia w3 data 
alexi_w3 <- read_csv("../../Wave_3/Data_scoring/alexithymia/alexithymia_scores_items_w3_20250523.csv")

# age at w3
demo3 <- read_csv("../../Wave_3/Data_scoring/demographics/demographics_clean_w3_20250527.csv") %>% 
  dplyr::select(
    participant_id,
    child_age_w3 = child_age_years_w3
  )

# primary caregiver education, income-to-needs relative to federal poverty line
demo_w1 <- read_csv("../../Wave_1/Data_scoring/demographics_parentproxy/demographics_clean_w1.csv") %>%
  dplyr::select(
    participant_id,
    federal_income_needs_primary_w1,
    primary_household_income_w1,
    primary_cg_education_w1
  )

# merge alexithymia w3 with rest of data
data <- data %>% dplyr::left_join(alexi_w3, by = "participant_id") %>% dplyr::left_join(demo3, by = "participant_id") %>% dplyr::left_join(demo_w1, by = "participant_id")

# read in new fci data 
fcinew <- read_csv("../../FCI_W1_W3_consolidation/fci_consolidated_final.csv") %>%
  dplyr::select(
    participant_id,
    num_transitions,
    age_left_bio_years,
    age_enter_stable_care_years,
    fci_timepoint
  ) %>%
  dplyr::rename(
    num_transitions_new = num_transitions,
    age_left_bio_years_new = age_left_bio_years,
    age_enter_stable_care_years_new = age_enter_stable_care_years
  )

data <- data %>%
  left_join(fcinew, by = "participant_id")

microfims <- read_csv("../../FIMS_microcoding/data/individual GSEQ yules q merged_validity_additional.csv") %>%
  dplyr::select(
    participant_id,
    pactive.cpositi,
    cactive.ppositi,
    pactive.cneutra,
    # possible covariates
    proportion_active_primary_parent,
    proportion_positive_primary_parent,
    proportion_active_primary_child,
    proportion_positive_primary_child
  )

# merge w old data
data <- data %>%
  left_join(microfims, by = "participant_id")

# descriptives: missing data rates, distributions, and associations w covariates for microfims
data %>% dplyr::count(!is.na(pactive.cpositi)) # N=143 have, 20 don't

# how many people have both alexi w1 and w3?
data %>% dplyr::count(!is.na(alex_idcom_mean_score_w3) & !is.na(alex_idcom_mean_score_w1)) # 72 with both 

ppositivcores <- read_csv("../../../../Jenn/caregiver FIMS positive CFAs/ppositivescores_negtask.csv") %>% dplyr::select(numid, ppositive, ppositive_se)

ppositivecommscores <- read_csv("../../../../Jenn/caregiver FIMS positive CFAs/ppositivecomscores_negtask.csv") %>% dplyr::select(numid, poscom, poscom_se)

fims_factors <- ppositivcores %>% full_join(ppositivecommscores, by = "numid")  

# add 0s to numid (001-009, 010-099, 100-163)
fims_factors <- fims_factors %>% 
  dplyr::mutate(
    numid = case_when(
      numid < 10 ~ str_c("00", as.character(numid), sep=""),
      numid >= 10 & numid < 100 ~ str_c("0", as.character(numid), sep=""),
      numid >= 100 ~ as.character(numid)
    )
  )

# now, add "MBB" to front of numid to make it into participant_id
fims_factors <- fims_factors %>% 
  dplyr::mutate(
    participant_id = str_c("MBB", numid, sep="")
  ) 

# merge factor scores w rest of data 
data <- data %>% left_join(fims_factors, by = "participant_id") %>% dplyr::select(-numid)
```

# Preliminary data cleaning
## Add variable for days between w1 and w3, years in stable care
```{r}
data <- data %>% 
  dplyr::mutate(
    years_between_w3_w1 = child_age_w3 - child_age_years_w1,
    days_since_w1_w3 = as.numeric(duration(years = years_between_w3_w1), "days")
  )

data %>% dplyr::count(!is.na(child_age_w3)) # only 76 people have age at w3

data <- data %>% 
  dplyr::mutate(
    years_in_stable_care = ifelse(ELA_group_w1==1, child_age_years_w1-age_enter_stable_care_years, child_age_years_w1),
    years_in_stable_care_new = ifelse(ELA_group_w1==1, child_age_years_w1-age_enter_stable_care_years_new, child_age_years_w1)
    )
```

## Add child race/ethnicity, in-person vs remote
```{r}
# add in child race and ethnicity data
cg_race <- read_csv("../../Wave_1/Data_scoring/demographics_parentproxy/demographics_clean_w1.csv") %>%
  dplyr::select(
    participant_id,
    primarycaregiver_race = primarycaregiver_race_w1,
    primarycaregiver_latinx = primarycaregiver_latinx_w1
  )

# read in list of in-person participants, make a variable to index in person
w1_inperson <- read_csv("../../Wave_1/Data_clean/wave_1_in_person_data_clean.csv") %>% dplyr::select(
  participant_id
) %>% dplyr::mutate(
  format = "inperson"
)

# merge with rest of the data
data <- data %>% left_join(cg_race, by = "participant_id") %>% left_join(w1_inperson, by = "participant_id")

# clean up the added vars
data <- data %>% dplyr::mutate(
  format = ifelse(is.na(format), "remote", format)
) %>% 
  dplyr::rename(child_race = child_race_w1)

# check race vars have been cleaned already -- yes
#unique(data$child_race)
#unique(data$primarycaregiver_race)

# make combined race/ethnicity 2 category variable for statistical analysis
data <- data %>% dplyr::mutate(
  child_race_ethnicity = case_when(
    child_race=="White_not_Latinx" & (child_latinx_w1==0 | is.na(child_latinx_w1)) ~ "White_not_Hispanic",
    is.na(child_race) & child_latinx_w1==0 ~ "White_not_Hispanic",
    child_race != "White_not_Latinx" | child_latinx_w1==1 ~ "POC_andor_Hispanic",
    is.na(child_race) & is.na(child_latinx_w1) ~ NA_character_
  ),
  primarycg_race_ethnicity = case_when(
    primarycaregiver_race=="White_not_Latinx" & (primarycaregiver_latinx==0 | is.na(primarycaregiver_latinx)) ~ "White_not_Hispanic",
    is.na(primarycaregiver_race) & primarycaregiver_latinx==0 ~ "White_not_Hispanic",
    primarycaregiver_race != "White_not_Latinx" | primarycaregiver_latinx==1 ~ "POC_andor_Hispanic",
    is.na(primarycaregiver_race) & is.na(primarycaregiver_latinx) ~ NA_character_
  )
)

data %>% dplyr::count(!is.na(child_race_ethnicity)) # 27 are missing child race ethnicity
data %>% dplyr::count(!is.na(child_race)) # 30 are missing race
data %>% dplyr::count(!is.na(child_latinx_w1)) # 34 are missing latinx
```

# Data exploration
## Variable distributions and outlier investigation
```{r}
primary_vars <- sample %>%
  dplyr::select(alex_idcom_mean_score_w1, ss_score_w1, poscom, ppositive, pactive.cpositi, years_in_stable_care_new, alex_idcom_mean_score_w3)

hist.data.frame(primary_vars)

# identify outliers-- boxplot method
identify_outliers(sample, alex_idcom_mean_score_w1) # no outliers
identify_outliers(sample, ss_score_w1) # no outliers
identify_outliers(sample, poscom) %>%
  dplyr::select(participant_id, ELA_group_w1, alex_idcom_mean_score_w1, poscom, is.outlier, is.extreme) #5 outliers 

poscom_outlier_ids <- identify_outliers(sample, poscom) %>%
  dplyr::select(participant_id)

identify_outliers(sample, ppositive)  %>%
  dplyr::select(participant_id, ELA_group_w1, ppositive, is.outlier, is.extreme) # 3 outliers

ppos_outlier_ids <- identify_outliers(sample, ppositive) %>%
  dplyr::select(participant_id)

identify_outliers(sample, pactive.cpositi) %>%
  dplyr::select(participant_id, ELA_group_w1, pactive.cpositi, is.outlier, is.extreme) # 2 outliers

conting_outlier_ids <- identify_outliers(sample, pactive.cpositi) %>%
  dplyr::select(participant_id)

identify_outliers(sample, years_in_stable_care_new) # no outliers

identify_outliers(sample, alex_idcom_mean_score_w3) #  no outliers
```

## Descriptive statistics
```{r}
# create dataset with manuscript sample: children who have baseline alexithymia data
sample <- data %>% 
  dplyr::filter(!is.na(alex_idcom_mean_score_w1))

# numbers in pre-, early-, mid-, late- adolescence in each wave
# pre: less than 10 years old; early: 10-12; mid: 13-15; late: 16-18
sample <- sample %>%
  mutate(
    adolescence_stage_w1 = case_when(
      child_age_years_w1 < 10 ~ "Pre-adolescence (<10)",
      child_age_years_w1 >= 10 & child_age_years_w1 < 13 ~ "Early adolescence (10–12)",
      child_age_years_w1 >= 13 & child_age_years_w1 < 16 ~ "Mid-adolescence (13–15)",
      child_age_years_w1 >= 16 & child_age_years_w1 <= 18 ~ "Late adolescence (16–18)",
      TRUE ~ NA_character_
    ),
    adolescence_stage_w3 = case_when(
      child_age_w3 < 10 ~ "Pre-adolescence (<10)",
      child_age_w3 >= 10 & child_age_w3 < 13 ~ "Early adolescence (10–12)",
      child_age_w3 >= 13 & child_age_w3 < 16 ~ "Mid-adolescence (13–15)",
      child_age_w3 >= 16 & child_age_w3 <= 18 ~ "Late adolescence (16–18)",
      TRUE ~ NA_character_
  )
  )

# number of families in sample
length(unique(sample$family_id))

# all demo variables -- full sample
demo <- sample %>%
  mutate(
    primary_cg_education_w1 = factor(primary_cg_education_w1, levels = c("highschool", "tech/vocational_school", "some_college", "associates degree", "bachelors_degree", "graduate_degree")),
    adolescence_stage_w1 = factor(adolescence_stage_w1, levels = c("Pre-adolescence (<10)", "Early adolescence (10–12)", "Mid-adolescence (13–15)", "Late adolescence (16–18)")),
    adolescence_stage_w3 = factor(adolescence_stage_w3, levels = c("Pre-adolescence (<10)", "Early adolescence (10–12)", "Mid-adolescence (13–15)", "Late adolescence (16–18)"))
  ) %>% 
  tbl_summary(include = c(ELA_group_w1, ela_3subgroup_str, format, child_race, child_race_ethnicity, child_sex_w1, primary_cg_education_w1, child_age_years_w1, primarycaregiver_race, federal_income_needs_primary_w1, adolescence_stage_w1, adolescence_stage_w3, alex_idcom_mean_score_w1, ss_score_w1, alex_idcom_mean_score_w3, ppositive, poscom, pactive.cpositi),
              statistic = list(
                all_continuous() ~ "{mean} ({sd}); Range: {min} - {max}"
    ))

table1a <- as_gt(demo) 
gtsave(table1a, filename = "figures/demos.docx")

# number missing caregiver-child interaction videos
sample %>% dplyr::count(!is.na(ppositive) & !is.na(poscom)) # 3 missing caregiver-child interaction data
sample %>% dplyr::count(!is.na(ss_score_w1)) # all complete
sample %>% dplyr::count(!is.na(years_in_stable_care_new)) # 10 missing

# demo variables -- by group
demo_groups <- sample %>%
  mutate(
    primary_cg_education_w1 = factor(primary_cg_education_w1, levels = c("highschool", "tech/vocational_school", "some_college", "associates degree", "bachelors_degree", "graduate_degree")),
    adolescence_stage_w1 = factor(adolescence_stage_w1, levels = c("Pre-adolescence (<10)", "Early adolescence (10–12)", "Mid-adolescence (13–15)", "Late adolescence (16–18)")),
    adolescence_stage_w3 = factor(adolescence_stage_w3, levels = c("Pre-adolescence (<10)", "Early adolescence (10–12)", "Mid-adolescence (13–15)", "Late adolescence (16–18)"))
  ) %>% 
  tbl_summary(by = ELA_group_w1, include = c(ELA_group_w1, ela_3subgroup_str, format, child_race, child_race_ethnicity, child_sex_w1, primary_cg_education_w1, child_age_years_w1, primarycaregiver_race, federal_income_needs_primary_w1, adolescence_stage_w1, adolescence_stage_w3, alex_idcom_mean_score_w1, alex_idcom_mean_score_w3, ss_score_w1, years_in_stable_care_new, ppositive, poscom, pactive.cpositi),
              statistic = list(
                all_continuous() ~ "{mean} ({sd}); Range: {min} - {max}"
    ))

table1b <- as_gt(demo_groups) 
gtsave(table1b, filename = "figures/demos_groups.docx")

# crEA group differences
t.test(child_age_years_w1 ~ ELA_group_w1, data = sample)

chisq.test(sample$child_sex_w1, sample$ELA_group_w1)

chisq.test(sample$child_race, sample$ELA_group_w1)

chisq.test(sample$primary_cg_education_w1, sample$ELA_group_w1)

chisq.test(sample$format, sample$ELA_group_w1)

chisq.test(sample$primarycaregiver_race, sample$ELA_group_w1)
tab <- table(sample$primarycaregiver_race, sample$ELA_group_w1)
pairwise.prop.test(tab, p.adjust.method = "bonferroni")

sample_ela <- sample %>% dplyr::filter(ELA_group_w1==1)
sample_comp <- sample %>% dplyr::filter(ELA_group_w1==0)

sample_ela %>% group_by(primarycaregiver_race) %>% summarise(n = n()) %>%
  mutate(freq = round(n / sum(n), digits=2)) %>% arrange(desc(n)) # 72% white

sample_comp %>% group_by(primarycaregiver_race) %>% summarise(n = n()) %>%
  mutate(freq = round(n / sum(n), digits=2)) %>% arrange(desc(n)) # 24% Asian

t.test(federal_income_needs_primary_w1 ~ ELA_group_w1, data = sample)

chisq.test(sample$tesi_lifetime, sample$ELA_group_w1)

t.test(leq_lifetime ~ ELA_group_w1, data = sample)

t.test(leq_recent_T1 ~ ELA_group_w1, data = sample)

t.test(tesi_recent_T1 ~ ELA_group_w1, data = sample)

sample %>% dplyr::count(!is.na(federal_income_needs_primary_w1)) # 100 out of 158 have this info
sample %>% dplyr::count(!is.na(primary_household_income_w1)) # 104 have income

t.test(alex_idcom_mean_score_w1 ~ ELA_group_w1, data=sample)
t.test(alex_idcom_mean_score_w3 ~ ELA_group_w1, data=sample)
t.test(ss_score_w1 ~ ELA_group_w1, data=sample)
t.test(ppositive ~ ELA_group_w1, data=sample)
t.test(poscom ~ ELA_group_w1, data=sample)
t.test(pactive.cpositi ~ ELA_group_w1, data=sample)

# format differences in primary study variables
t.test(alex_idcom_mean_score_w1 ~ format, data=sample)
t.test(alex_idcom_mean_score_w3 ~ format, data=sample)
t.test(ss_score_w1 ~ format, data=sample)
t.test(ppositive ~ format, data=sample) # sig
t.test(poscom ~ format, data=sample) # sig
t.test(pactive.cpositi ~ format, data=sample)
```

## Differences in drop-out according to demographics
```{r}
# check
sample %>% dplyr::count(is.na(alex_idcom_pct_complete_w3))

# create dropped out variable
sample <- sample %>%
  mutate(
    droppedout = ifelse(is.na(alex_idcom_pct_complete_w3), 1, 0)
  )

sample_ela <- sample_ela  %>%
  mutate(
    droppedout = ifelse(is.na(alex_idcom_pct_complete_w3), 1, 0)
  )

sample_comp <- sample_comp  %>%
  mutate(
    droppedout = ifelse(is.na(alex_idcom_pct_complete_w3), 1, 0)
  )

# group, child sex, race, primary caregiver education, primary caregiver race, income-to-needs
chisq.test(sample$ELA_group_w1, sample$droppedout) # ns
chisq.test(sample$child_sex_w1, sample$droppedout) # ns
chisq.test(sample$child_race_ethnicity, sample$droppedout) # ns
chisq.test(sample$primary_cg_education_w1, sample$droppedout) #ns
chisq.test(sample$primarycg_race_ethnicity, sample$droppedout) # ns
t.test(federal_income_needs_primary_w1 ~ droppedout, data = sample) # ns

# variables of interst
t.test(alex_idcom_mean_score_w1 ~ droppedout, data = sample) # ns
t.test(ss_score_w1 ~ droppedout, data = sample) # ns
t.test(poscom ~ droppedout, data = sample) # ns
t.test(ppositive ~ droppedout, data = sample) # ns (p = .055)
t.test(pactive.cpositi ~ droppedout, data = sample) # ns

# variables of interest by group -- ela first
t.test(alex_idcom_mean_score_w1 ~ droppedout, data = sample_ela) # ns
t.test(ss_score_w1 ~ droppedout, data = sample_ela) # ns
t.test(poscom ~ droppedout, data = sample_ela) # ns
t.test(ppositive ~ droppedout, data = sample_ela) # ns 
t.test(pactive.cpositi ~ droppedout, data = sample_ela) # ns
t.test(years_in_stable_care_new ~ droppedout, data=sample_ela)

# comp group
t.test(alex_idcom_mean_score_w1 ~ droppedout, data = sample_comp) # ns
t.test(ss_score_w1 ~ droppedout, data = sample_comp) # ns
t.test(poscom ~ droppedout, data = sample_comp) # ns
t.test(ppositive ~ droppedout, data = sample_comp) # ns 
t.test(pactive.cpositi ~ droppedout, data = sample_comp) # ns

# recollected fci vs w1 fci
sample %>% dplyr::count(fci_timepoint.y)

# differences in years in stable care based on reporting timepoint?
chisq.test(sample$fci_timepoint.y, sample$years_in_stable_care_new) # ns
chisq.test(sample$fci_timepoint.y, sample$num_transitions_new) # ns
```

### Table: differences in primary study variables based on drop-out
```{r}
# Variables to test
vars <- c("alex_idcom_mean_score_w1", "ss_score_w1", "poscom", "ppositive", "pactive.cpositi") 

# Function to run t-tests and format results
get_ttest_results <- function(data, vars, group_var = "droppedout") {
  sapply(vars, function(v) {
    
    # Formula for t.test
    formula <- as.formula(paste(v, "~", group_var))
    
    # Run t-test
    tt <- t.test(formula, data = data)
    
    # Extract values
    t_val <- unname(tt$statistic)
    df <- unname(tt$parameter)
    p <- tt$p.value
    
    # Format p-value
    p_text <- if (p < .001) {
      "< .001"
    } else {
      sprintf("= %.3f", p)
    }
    
    # Create formatted string
    sprintf("t(%.1f) = %.2f, p %s",
            df,
            t_val,
            p_text)
  })
}

# Create the summary table
results_table <- data.frame(
  Variable = vars,
  full_sample = get_ttest_results(sample, vars),
  crea_group = get_ttest_results(sample_ela, vars),
  comp_group = get_ttest_results(sample_comp, vars),
  row.names = NULL,
  check.names = FALSE
)

results_table

write_csv(results_table, "figures/tables1_dropout.csv")
```


## Reliability
Alexithymia, security scale
```{r}
# tas: 
# ident: 1, 3, 6, 7, 9, 13, 14
# desc: 2, 4, 11, 12, 17
# our alexithymia: 
# ident: 1, 2, 3, 4, 6, 7
# desc: 8, 9, 10, 11, 12, 

# alexithymia w1
alex_ident_items_w1 <- sample %>%
  dplyr::select(alexithymia_1_w1:alexithymia_7_w1) %>%
  drop_na()

ltm::cronbach.alpha(alex_ident_items_w1, standardized=TRUE)

alex_desc_items_w1 <- sample %>%
  dplyr::select(alexithymia_8_w1:alexithymia_12_w1) %>%
  drop_na()

ltm::cronbach.alpha(alex_desc_items_w1, standardized=TRUE)

# alexithymia w3
alex_ident_items_w3 <- sample %>%
  dplyr::select(alexithymia_1_w3:alexithymia_7_w3) %>%
  drop_na()

ltm::cronbach.alpha(alex_ident_items_w3, standardized=TRUE)

alex_desc_items_w3 <- sample %>%
  dplyr::select(alexithymia_8_w3:alexithymia_12_w3) %>%
  drop_na()

ltm::cronbach.alpha(alex_desc_items_w3, standardized=TRUE)

# security scale
ss_items <- sample %>%
  dplyr::select(ss_1_score:ss_15_score) %>%
  drop_na()

ltm::cronbach.alpha(ss_items, standardized=TRUE)
```

## Missing data item level
```{r}
# security scale
sample %>% dplyr::count(missing_count)

# alexithymia baseline
sample %>% dplyr::count(alex_ident_pct_complete_w1) # 7 items (6 out of 7 = 0.857; 5 out of 7 = 0.71)
sample %>% dplyr::count(alex_comm_pct_complete_w1) # 5 items (80% = 4/5)

sample <- sample %>%
  dplyr::mutate(
    alex_idcom_pct_complete_w1 = (alex_ident_pct_complete_w1 + alex_comm_pct_complete_w1) /2
  )

sample %>% dplyr::count(alex_idcom_pct_complete_w1) # 11/12, 10/12, 9/12

# alexithymia followup
sample %>% dplyr::count(alex_idcom_pct_complete_w3) # 12 items (0.91 = 11/12)
```


## Bivariate correlations table
Years in stable care, Alexithymia W1, Alexithymia W3, Security Scale, poscom, ppositive, pactive.cpositi
```{r}
cont_vars <- sample %>% dplyr::select(ppositive, poscom, pactive.cpositi, years_in_stable_care_new, alex_idcom_mean_score_w1, alex_idcom_mean_score_w3, ss_score_w1) %>% drop_na()

testRes = cor.mtest(cont_vars, conf.level = 0.95)
M = cor(cont_vars)
corrplot(M, p.mat = testRes$p, sig.level = 0.05, order = 'hclust', addrect = 2) # physical contact, supportive, positive/warm, and warmth are negatively correlated with age

corrplot(M,
         method = "number",   # show correlation values only
         type = "upper",
         order = "hclust",
         addrect = 2)

# years in stable care correlations -- ELA group only
cor.test(sample_ela$alex_idcom_mean_score_w1, sample_ela$years_in_stable_care_new)
cor.test(sample_ela$alex_idcom_mean_score_w3, sample_ela$years_in_stable_care_new)
cor.test(sample_ela$ss_score_w1, sample_ela$years_in_stable_care_new)
cor.test(sample_ela$ppositive, sample_ela$years_in_stable_care_new)
cor.test(sample_ela$poscom, sample_ela$years_in_stable_care_new)
cor.test(sample_ela$pactive.cpositi, sample_ela$years_in_stable_care_new)

# age and alexithymia at baseline in crEA group only
cor.test(sample_ela$alex_idcom_mean_score_w1, sample_ela$child_age_years_w1)

# get other p-values for chart
cor.test(sample$alex_idcom_mean_score_w1, sample$alex_idcom_mean_score_w3)
cor.test(sample$alex_idcom_mean_score_w1, sample$ss_score_w1)
cor.test(sample$alex_idcom_mean_score_w3, sample$ss_score_w1)

desc <- sample %>%
  tbl_summary(include = c(ppositive, poscom, pactive.cpositi, years_in_stable_care, alex_idcom_mean_score_w1, alex_idcom_mean_score_w3, ss_score_w1),
              statistic = list(
                all_continuous() ~ "{mean} ({sd})"
    ))

table2 <- as_gt(desc) 
gtsave(table2, filename = "figures/desc_stats.docx")

mean(sample_ela$years_in_stable_care, na.rm=TRUE) #7.72
sd(sample_ela$years_in_stable_care, na.rm=TRUE) #4.41
```

## Covariate selection: Alexithymia associations with race, sex, inperson vs. remote, age
```{r}
# alexithymia and age correlation
cor.test(data$alex_idcom_mean_score_w1, data$child_age_years_w1, method = c("pearson")) #ns
cor.test(data$alex_idcom_mean_score_w3, data$child_age_w3, method = c("pearson")) #ns

# alexithymia t-test with sex, race/eth, and inperson vs. remote
t.test(alex_idcom_mean_score_w1 ~ child_sex_w1, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) #ns (p=.055), females higher
t.test(alex_idcom_mean_score_w3 ~ child_sex_w1, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) #ns

t.test(alex_idcom_mean_score_w1 ~ child_race_ethnicity, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) #sig (p=.044), white higher
t.test(alex_idcom_mean_score_w3 ~ child_race_ethnicity, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) #sig (p = .047), white higher

t.test(alex_idcom_mean_score_w1 ~ format, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) #ns

# alexithymia anova with race
summary(aov(alex_idcom_mean_score_w1 ~ child_race, data=data)) #ns
alex_race_w3 <- aov(alex_idcom_mean_score_w3 ~ child_race, data=data)
summary(alex_race_w3)# sig, p = .025
TukeyHSD(alex_race_w3) # no sig pairwise differences

data %>% dplyr::group_by(child_race) %>% dplyr::summarize(mean = mean(alex_idcom_mean_score_w3, na.rm=TRUE), n=n()) # black and Native american means are higher, but there are only 4 black children and 2 native

data %>% dplyr::count(!is.na(alex_idcom_mean_score_w3) & child_race=="Black/AA") # only 2 black children have alexithymia in W3

ggplot(data, aes(x = child_age_years_w1)) +
  geom_histogram()

data %>% dplyr::count(child_age_years_w1 < 10) # ~1/2 of the sample is younger than 10 at W1
data %>% dplyr::count(child_age_w3 < 10) # but only 12 out of 76 are younger than 10 at W3
```

# Covariate selection: macroFIMS associations with race, sex, inperson vs. remote, age
factors - pos/warm and social communication from *negative* task
freq_pos_affect_mother_master_neg, intensity_freq_pos_affect_mother_master_neg, intensity_pos_affect_mother_master_neg, warmth_mother_master_neg, humor_mother_master_neg, physical_contact_mother_master_neg, supportive_mother_master_neg

Jenn's writeup of factor analysis: Users/Jenn/caregiver FIMS positive CFAs
```{r}
# read in factor scores from Jenn's mplus analysis
ppositivcores <- read_csv("../../../../Jenn/caregiver FIMS positive CFAs/ppositivescores_negtask.csv") %>% dplyr::select(numid, ppositive, ppositive_se)

# "../../../../../Users/Jenn/caregiver FIMS positive CFAs/ppositivescores_negtask.csv"

ppositivecommscores <- read_csv("../../../../Jenn/caregiver FIMS positive CFAs/ppositivecomscores_negtask.csv") %>% dplyr::select(numid, poscom, poscom_se)

poscomscores_postask <- read_csv("../../../../Jenn/caregiver FIMS positive CFAs/poscomscores_positivetask_final.csv") %>% 
  dplyr::select(numid, poscom, poscom_se) %>% 
  dplyr::rename(poscom_postask = poscom,
                poscom_se_postask = poscom_se)

ppositive_postask <- read_csv("../../../../Jenn/caregiver FIMS positive CFAs/ppositivescores_postask_final.csv") %>% 
  dplyr::select(numid, ppositive, ppositive_se) %>%
  dplyr::rename(ppositive_postask = ppositive,
                ppositive_se_postaxk = ppositive_se)

fims_factors <- ppositivcores %>% full_join(ppositivecommscores, by = "numid")  %>% full_join(poscomscores_postask, by = "numid")  %>% full_join(ppositive_postask, by = "numid")

# add 0s to numid (001-009, 010-099, 100-163)
fims_factors <- fims_factors %>% 
  dplyr::mutate(
    numid = case_when(
      numid < 10 ~ str_c("00", as.character(numid), sep=""),
      numid >= 10 & numid < 100 ~ str_c("0", as.character(numid), sep=""),
      numid >= 100 ~ as.character(numid)
    )
  )

# now, add "MBB" to front of numid to make it into participant_id
fims_factors <- fims_factors %>% 
  dplyr::mutate(
    participant_id = str_c("MBB", numid, sep="")
  ) 

# merge factor scores w rest of data 
data <- data %>% left_join(fims_factors, by = "participant_id") %>% dplyr::select(-numid)

# fims t-tests with sex, race/eth, and inperson vs. remote
t.test(ppositive ~ child_sex_w1, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) #ns
t.test(poscom ~ child_sex_w1, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) #ns

t.test(ppositive ~ primarycg_race_ethnicity, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) #ns
t.test(poscom ~ primarycg_race_ethnicity, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) # POC and/or hispanic lower (p=.004)

t.test(ppositive ~ format, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) # in-person higher (p=.01)
t.test(poscom ~ format, data = data,
       alternative = c("two.sided"), mu=0, var.equal = FALSE,
       conf.level = 0.95, paired = FALSE) # in-person lower (p=.002)
```
Note: due to the differences by race/eth and format, we will add those as covariates to macroFIMS analyses

# Alexithymia associations with child behavior in the tasks
```{r}
# alexithymia, child age, child neg affect, and child pos affect in correlation matrix
child_affect <- data %>% dplyr::select(ppositive, poscom, childpostotal_negative, childnegtotal_negative, alex_idcom_mean_score_w1, child_age_years_w1) %>% drop_na()

testRes = cor.mtest(child_affect, conf.level = 0.95)
M = cor(child_affect)
corrplot(M, p.mat = testRes$p, sig.level = 0.05, order = 'hclust', addrect = 2) # alexithymia and ppositive positively related to child positive affect in neg interaction, child negative affect negatively related to ppositive

cor.test(data$alex_idcom_mean_score_w1, data$childpostotal_negative, method= c("pearson")) #r = .18, p = .026

# associations w affect controlling for child age and sex
child_pos_affect <- lm(alex_idcom_mean_score_w1 ~ childpostotal_negative + child_age_years_w1 + child_sex_w1, data = data)
tab_model(child_pos_affect, vcov.fun = "CL", vcov.args = list(type="HC3", cluster = data$family_id), show.se=TRUE) # ns (p = 0.06)

child_neg_affect <- lm(alex_idcom_mean_score_w1 ~ childnegtotal_negative + child_age_years_w1 + child_sex_w1, data = data)
tab_model(child_neg_affect, vcov.fun = "CL", vcov.args = list(type="HC3", cluster = data$family_id), show.se=TRUE) #ns
```

# Aim 1: Alexithymia differences across Study Groups

## Latent Growth Models: Baseline and Change Over Time Differences
covariates: sex, age
```{r}
# LMM -- complete cases only
# make a long format dataset with alexithymia and timepoint, security score, age at w1, sex
ss_graph_long <- data %>%
  dplyr::select(participant_id, ss_score_w1, alex_idcom_mean_score_w1, alex_idcom_mean_score_w3, child_sex_w1, child_age_years_w1, child_race_eth_bin, ELA_group_w1) %>%
  pivot_longer(
    cols = c(alex_idcom_mean_score_w1, alex_idcom_mean_score_w3),
    names_to = "time",
    values_to = "alex_idcom_mean_score"
  ) %>% dplyr::mutate(
    time = ifelse(time=="alex_idcom_mean_score_w1", "w1", "w3"),
    time = factor(time, levels = c("w1", "w3"))
  )

# center variables
ss_graph_long$ss_c <- scale(ss_graph_long$ss_score_w1, center = TRUE, scale = FALSE)[,1]
ss_graph_long$age_c  <- scale(ss_graph_long$child_age_years_w1, center = TRUE, scale = FALSE)[,1]
ss_graph_long$female_c <- scale(ifelse(ss_graph_long$child_sex_w1 == "female", 1, 0), center = TRUE, scale = FALSE)[,1]
ss_graph_long$race_c <- scale(ss_graph_long$child_race_eth_bin, center = TRUE, scale = FALSE)[,1]
ss_graph_long$ELA_group_c <- scale(ss_graph_long$ELA_group_w1, center = TRUE, scale = FALSE)[,1]

groupchange_model <- lmerTest::lmer(
  alex_idcom_mean_score ~ time * ELA_group_c + age_c + female_c + (1 | participant_id),
  data = ss_graph_long
)

summary(groupchange_model)

# latent growth model: FIML for missing data
model_conditional <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ ELA_group_w1 + child_age_years_w1 + child_sex_w1
  s ~ ELA_group_w1 + child_age_years_w1 + child_sex_w1

  # Covariance between growth factors (residual after prediction)
  i ~~ s'

fit_conditional <- growth(model_conditional,
                          data = data,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_conditional, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)

# are the means different at T2?
# center age and sex
data$age_c  <- scale(data$child_age_years_w1, center = TRUE, scale = FALSE)[,1]
data$female_c <- scale(ifelse(data$child_sex_w1 == "female", 1, 0), center = TRUE, scale = FALSE)[,1]

model_followup <- '
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  i ~ b_cg*ELA_group_w1 + b_age*age_c + b_sex*female_c
  s ~ g0*1 + g_cg*ELA_group_w1 + g_age*age_c + g_sex*female_c

  i ~~ s
  
  # define mean intercept and slope for control group
  i ~ i_mean*1
  
  # Define predicted T2 means for each group
  # (at mean levels of covariates, i.e. age = 0, sex = 0 after centering)
  # T2 = intercept + 1*slope  (because slope loading at T2 = 1)

  T2_normative := i_mean + g0                        # caregiving = 0
  T2_adverse   := (i_mean + b_cg) + (g0 + g_cg)     # caregiving = 1

  # Difference between groups at T2
  T2_diff := T2_adverse - T2_normative
  
  # define each slope for each group
  slope_normative := g0 + g_cg*0
  slope_adverse := g0 + g_cg*1
'

fit_followup <- growth(model_followup,
                       data = data,
                       missing = "fiml",
                       estimator = "MLR",
                       cluster = "family_id")

summary(fit_followup, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE, ci = TRUE)
standardizedSolution(fit_followup, type = "std.all", ci=TRUE, level = .95)
```

# Aim 2: Caregiver-Child Interaction Dynamics and Alexithymia

## 2a: Time in Stable Care and Alexithymia
### Correlation between time in stable care and number of transitions
```{r}
cor.test(sample_ela$years_in_stable_care_new, sample_ela$num_transitions_new) # more time in stable care -- fewer transitions
```

### Baseline differences only
```{r}
stablecare_transitions_alex_w1 <- lm(alex_idcom_mean_score_w1 ~ years_in_stable_care_new + num_transitions_new + child_age_years_w1 + child_sex_w1, data = sample_ela)
tab_model(stablecare_transitions_alex_w1, show.std= "std", vcov.fun = "HC3", vcov.args = list(cluster = data_ela$family_id), show.se = TRUE) # years in stable care is significant; B = -0.48, p = .026
# num transitions is marginal B = -0.29, p = .054

# collinearity test
library(car)
vif(stablecare_transitions_alex_w1) # all lower than 3 

model_timeinstablecare <- '
  # Caregiving and covariates predicting growth factors
  alex_idcom_mean_score_w1 ~ years_in_stable_care_new + num_transitions_new + child_age_years_w1 + child_sex_w1
'

fit_model_timeinstablecare <- sem(model_timeinstablecare,
                              data = sample_ela,
                              missing = "fiml",
                              estimator = "MLR",
                              cluster = "family_id")

summary(fit_model_timeinstablecare, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_model_timeinstablecare, type = "std.all", ci=TRUE, level = .95)
```

## 2b: Attachment Security and Alexithymia
### Latent Growth Model: Baseline and Change over time differences
```{r}
# complete cases only
# make a long format dataset with alexithymia and timepoint, security score, age at w1, sex
ss_graph_long <- data %>%
  dplyr::select(participant_id, ss_score_w1, alex_idcom_mean_score_w1, alex_idcom_mean_score_w3, child_sex_w1, child_age_years_w1, child_race_eth_bin, ELA_group_w1) %>%
  pivot_longer(
    cols = c(alex_idcom_mean_score_w1, alex_idcom_mean_score_w3),
    names_to = "time",
    values_to = "alex_idcom_mean_score"
  ) %>% dplyr::mutate(
    time = ifelse(time=="alex_idcom_mean_score_w1", "w1", "w3"),
    time = factor(time, levels = c("w1", "w3"))
  )

# center variables
ss_graph_long$ss_c <- scale(ss_graph_long$ss_score_w1, center = TRUE, scale = FALSE)[,1]
ss_graph_long$age_c  <- scale(ss_graph_long$child_age_years_w1, center = TRUE, scale = FALSE)[,1]
ss_graph_long$female_c <- scale(ifelse(ss_graph_long$child_sex_w1 == "female", 1, 0), center = TRUE, scale = FALSE)[,1]
ss_graph_long$race_c <- scale(ss_graph_long$child_race_eth_bin, center = TRUE, scale = FALSE)[,1]
ss_graph_long$ELA_group_c <- scale(ss_graph_long$ELA_group_w1, center = TRUE, scale = FALSE)[,1]

sschange_model <- lmerTest::lmer(
  alex_idcom_mean_score ~ time * ss_c + ELA_group_c + age_c + female_c + (1 | participant_id),
  data = ss_graph_long
)

summary(sschange_model)

mean_ss <- mean(sample$ss_score_w1, na.rm=TRUE) # mean is 3.10
sd(sample$ss_score_w1, na.rm=TRUE) # sd is 0.54
sd1up_ss <- mean(sample$ss_score_w1, na.rm=TRUE) + sd(sample$ss_score_w1, na.rm=TRUE) # 3.65
sd1down_ss <- mean(sample$ss_score_w1, na.rm=TRUE) - sd(sample$ss_score_w1, na.rm=TRUE) # 0.92

# growth model with FIML for missing data
model_ss_conditional <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ bo*1 + b_ss*ss_score_w1 + b_group*ELA_group_w1 + b_age*child_age_years_w1 + b_sex*female + b_race*primarycg_poc
  s ~ g1*1 + g_ss*ss_score_w1  + g_group*ELA_group_w1 + g_age*child_age_years_w1 + g_sex*female + g_race*primarycg_poc

  # Covariance between growth factors (residual after prediction)
  i ~~ s

  # Predicted growth slope at each level of ss (holding age and sex at mean) 
  slope_ss_low  := g1 + g_ss*0.92  + g_age*11.28 + g_sex*0.49 + g_race*0.5
  slope_ss_mean := g1 + g_ss*3.10 + g_age*11.28 + g_sex*0.49 + g_race*0.5
  slope_ss_high := g1 + g_ss*3.65 + g_age*11.28 + g_sex*0.49 + g_race*0.5'

fit_ss_conditional <- growth(model_ss_conditional,
                          data = sample,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_ss_conditional, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_ss_conditional, type = "std.all", ci=TRUE, level = .95)
```

## 2c: MacroFIMS and Alexithymia
### Positive/Warm Parenting: Latent Growth Model: Baseline and Change over time differences
```{r}
sample <- sample %>%
  mutate(female = ifelse(child_sex_w1 == "female", 1, 0),
         primarycg_poc = ifelse(primarycg_race_ethnicity == "POC_andor_Hispanic", 1, 0))

mean(sample$child_age_years_w1, na.rm=TRUE) #11.28
mean(sample$female, na.rm=TRUE) # 0.49
mean(sample$primarycg_poc, na.rm=TRUE) # 0.5

model_ppositive <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ bo*1 + b_pp*ppositive + b_group*ELA_group_w1 + b_age*child_age_years_w1 + b_sex*female + b_race*primarycg_poc + b_format*format
  s ~ g1*1 + g_pp*ppositive  + g_group*ELA_group_w1 + g_age*child_age_years_w1 + g_sex*female + g_race*primarycg_poc + g_format*format

  # Covariance between growth factors (residual after prediction)
  i ~~ s

  # Predicted growth slope at each level of pppositive (holding age and sex at mean) 
  slope_pp_low  := g1 + g_pp*-0.945  + g_age*11.28 + g_sex*0.49 + g_race*0.5
  slope_pp_mean := g1 + g_pp*-0.011 + g_age*11.28 + g_sex*0.49 + g_race*0.5
  slope_pp_high := g1 + g_pp*0.921 + g_age*11.28 + g_sex*0.49 + g_race*0.5'
  
fit_ppositive <- growth(model_ppositive,
                          data = sample,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_ppositive, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_ppositive, type = "std.all", ci=TRUE, level = .95)

fscores_ppositive <- lavPredict(fit_ppositive, type = "lv")

# run model without ppostivie outliers
sample_noppos_out <- sample %>%
  dplyr::filter(!(participant_id %in% ppos_outlier_ids$participant_id))


fit_ppositive_noout <- growth(model_ppositive,
                          data = sample_noppos_out,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")
summary(fit_ppositive_noout, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE) # no longer effect

# graph effect on slope to visualize
# make a long format dataset with alexithymia and timepoint, ppositive, age at w1, sex, race/ethnicity
ppositive_graph_long <- data %>%
  dplyr::select(participant_id, ppositive, alex_idcom_mean_score_w1, alex_idcom_mean_score_w3, child_sex_w1, child_age_years_w1, primarycg_race_ethnicity, ELA_group_w1) %>%
  pivot_longer(
    cols = c(alex_idcom_mean_score_w1, alex_idcom_mean_score_w3),
    names_to = "time",
    values_to = "alex_idcom_mean_score"
  ) %>% dplyr::mutate(
    time = ifelse(time=="alex_idcom_mean_score_w1", "w1", "w3"),
    time = factor(time, levels = c("w1", "w3")),
    cg_race_bin = ifelse(primarycg_race_ethnicity == "White_not_Hispanic", 0, 1)
  )

# center variables
ppositive_graph_long$ppositive_c <- scale(ppositive_graph_long$ppositive, center = TRUE, scale = FALSE)[,1]
ppositive_graph_long$age_c  <- scale(ppositive_graph_long$child_age_years_w1, center = TRUE, scale = FALSE)[,1]
ppositive_graph_long$female_c <- scale(ifelse(ppositive_graph_long$child_sex_w1 == "female", 1, 0), center = TRUE, scale = FALSE)[,1]
ppositive_graph_long$cgrace_c <- scale(ppositive_graph_long$cg_race_bin, center = TRUE, scale = FALSE)[,1]
ppositive_graph_long$ELA_group_c <- scale(ppositive_graph_long$ELA_group_w1, center = TRUE, scale = FALSE)[,1]

# specify model with compelte data
ppositivechange_model <- lmerTest::lmer(
  alex_idcom_mean_score ~ time:ppositive_c + time + ppositive_c + ELA_group_c + age_c + female_c + cgrace_c  + (1 | participant_id),
  data = ppositive_graph_long
)

summary(ppositivechange_model)
emmeans::emtrends(ppositivechange_model, ~time, var = "ppositive_c")

library(effects)

mean_pp <- mean(sample$ppositive, na.rm=TRUE) # mean is -0.011
sd(sample$ppositive, na.rm=TRUE) # sd is 23.48
sd1up <- mean(sample$ppositive, na.rm=TRUE) + sd(sample$ppositive, na.rm=TRUE) # 89.73
sd1down <- mean(sample$ppositive, na.rm=TRUE) - sd(sample$ppositive, na.rm=TRUE) # 42.767

# Generate effect data
effects_ppositive <- effect(
  term = "time:ppositive_c",
  mod = ppositivechange_model,
  xlevels = list(
    timepoint0 = c('w1', 'w3'),
    ppositive_c = c(sd1down, sd1up)  # -1 SD and +1 SD
  )
)

effectsdata_ppositive <- as.data.frame(effects_ppositive)

effectsdata_ppositive$ppositive_c <- factor(
  effectsdata_ppositive$ppositive_c,
  levels = c(sd1down, sd1up),
  labels = c("-1SD parenting factor", "+1SD parenting factor")
)

emmeans::emmeans(ppositivechange_model, ~ time | ppositive_c, at = list(ppositive_c = c(sd1down, mean_pp, sd1up))) # significant decrease across time at noderate and high levels of positive/warm parenting, n.s. for low
# to get the slopes and 95% CIs, subtract w1 from w3 values within each of the values of ppositive

ppositive_graph_long <- ppositive_graph_long %>%
  mutate(ELA_group_w1_f = factor(ELA_group_w1, 
                                levels = c(0, 1), 
                                labels = c("Comparison", "crEA-Exposed")))

#yellow and green colors
dark2_cols <- RColorBrewer::brewer.pal(8, "Dark2")[c(5, 6)]

# Plot with both effect predictions and raw datapoints
ppos_plot <- ggplot() +
  #geom_point(data = ppositive_graph_long, aes(x = time, y = alex_idcom_mean_score, group = participant_id,
  #color = ELA_group_w1_f), alpha = 0.7, position = position_jitter(width = 0.05, height = 0)) +
  #geom_line(data = ppositive_graph_long, aes(x = time, y = alex_idcom_mean_score, color = ELA_group_w1_f, group = participant_id), size = 1, alpha=0.5) +
  # Predicted effects
  geom_point(data = effectsdata_ppositive, aes(x = time, y = fit, group = ppositive_c), size = 3) +
  geom_line(data = effectsdata_ppositive, aes(x = time, y = fit, group = ppositive_c, linetype = ppositive_c), size = 1) +
  scale_linetype_manual(
    name = "Simple Slopes",
    values = c("solid", "dotted")
  ) +
  xlab("Timepoint") + 
  scale_x_discrete(
  breaks = c('w1', 'w3'),
  labels = c("Baseline", "Followup"),
  expand = c(0, 0.3)
) + 
  ylab("Alexithymia") +
  labs(color = "crEA-Exposure Group") + 
  #scale_color_brewer(palette = "Spectral") +
  scale_color_manual(
    name = "crEA-Exposure Group",
    values = c("Comparison" = dark2_cols[1], "crEA-Exposed" = dark2_cols[2])
  ) +
  ggtitle("Positive/Warm Parenting") +
  theme_bw() + theme(panel.grid.major = element_blank(),
                         panel.grid.minor = element_blank(),
                         aspect.ratio = .6,
                         panel.border = element_rect(
                           color = "black",  
                           size = 0.5),
                         axis.title = element_text(size = 13),
                         axis.text = element_text(size = 11),
                        plot.title = element_text(hjust = 0.5, face="bold"),
                        legend.position = "bottom")
```

### Poscom: Latent Growth Model: Baseline and Change over time differences
```{r}
model_poscom <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ bo*1 + b_pp*poscom + b_group*ELA_group_w1 + b_age*child_age_years_w1 + b_sex*female + b_race*primarycg_poc + b_format*format
  s ~ g1*1 + g_pp*poscom  + g_group*ELA_group_w1 + g_age*child_age_years_w1 + g_sex*female + g_race*primarycg_poc + g_format*format

  # Covariance between growth factors (residual after prediction)
  i ~~ s
  '
  
fit_poscom <- growth(model_poscom,
                          data = sample,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_poscom, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_poscom, type = "std.all", ci=TRUE, level = .95)

# run model without poscom outliers
sample_noposcom_out <- sample %>%
  dplyr::filter(!(participant_id %in% poscom_outlier_ids$participant_id))


fit_poscom_noout <- growth(model_poscom,
                          data = sample_noposcom_out,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")
summary(fit_poscom_noout, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE) # no longer effect on baseline alex
standardizedSolution(fit_poscom_noout, type = "std.all", ci=TRUE, level = .95)

# graph effect on slope to visualize
# make a long format dataset with alexithymia and timepoint, ppositive, age at w1, sex, race/ethnicity
poscom_graph_long <- data %>%
  dplyr::select(participant_id, poscom, alex_idcom_mean_score_w1, alex_idcom_mean_score_w3, child_sex_w1, child_age_years_w1, primarycg_race_ethnicity, ELA_group_w1) %>%
  pivot_longer(
    cols = c(alex_idcom_mean_score_w1, alex_idcom_mean_score_w3),
    names_to = "time",
    values_to = "alex_idcom_mean_score"
  ) %>% dplyr::mutate(
    time = ifelse(time=="alex_idcom_mean_score_w1", "w1", "w3"),
    time = factor(time, levels = c("w1", "w3")),
    cg_race_bin = ifelse(primarycg_race_ethnicity == "White_not_Hispanic", 0, 1)
  )

# center variables
poscom_graph_long$poscom_c <- scale(poscom_graph_long$poscom, center = TRUE, scale = FALSE)[,1]
poscom_graph_long$age_c  <- scale(poscom_graph_long$child_age_years_w1, center = TRUE, scale = FALSE)[,1]
poscom_graph_long$female_c <- scale(ifelse(poscom_graph_long$child_sex_w1 == "female", 1, 0), center = TRUE, scale = FALSE)[,1]
poscom_graph_long$cgrace_c <- scale(poscom_graph_long$cg_race_bin, center = TRUE, scale = FALSE)[,1]
poscom_graph_long$ELA_group_c <- scale(poscom_graph_long$ELA_group_w1, center = TRUE, scale = FALSE)[,1]

# specify model with compelte data
poscomchange_model <- lmerTest::lmer(
  alex_idcom_mean_score ~ time:poscom_c + time + poscom_c + ELA_group_c + age_c + female_c + cgrace_c  + (1 | participant_id),
  data = poscom_graph_long
)

summary(poscomchange_model)
emmeans::emtrends(poscomchange_model, ~time, var = "poscom_c")

library(effects)

mean_poscom <- mean(sample$poscom, na.rm=TRUE) # mean is -0.011
sd(sample$poscom, na.rm=TRUE) # sd is 23.48
sd1up_pos <- mean(sample$poscom, na.rm=TRUE) + sd(sample$poscom, na.rm=TRUE) # 89.73
sd1down_posc <- mean(sample$poscom, na.rm=TRUE) - sd(sample$poscom, na.rm=TRUE) # 42.767

# Generate effect data
effects_poscom <- effect(
  term = "time:poscom_c",
  mod = poscomchange_model,
  xlevels = list(
    timepoint0 = c('w1', 'w3'),
    poscom_c = c(sd1down_posc, sd1up_pos)  # -1 SD and +1 SD
  )
)

effects_poscom <- as.data.frame(effects_poscom)

effects_poscom$poscom_c <- factor(
  effects_poscom$poscom_c,
  levels = c(sd1down_posc, sd1up_pos),
  labels = c("-1SD parenting factor", "+1SD parenting factor")
)

#emmeans::emmeans(ppositivechange_model, ~ time | ppositive_c, at = list(ppositive_c = c(sd1down, mean_pp, sd1up))) # significant decrease across time at noderate and high levels of positive/warm parenting, n.s. for low
# to get the slopes and 95% CIs, subtract w1 from w3 values within each of the values of ppositive

poscom_graph_long <- poscom_graph_long %>%
  mutate(ELA_group_w1_f = factor(ELA_group_w1, 
                                levels = c(0, 1), 
                                labels = c("Comparison", "crEA-Exposed")))

#yellow and green colors
dark2_cols <- RColorBrewer::brewer.pal(8, "Dark2")[c(5, 6)]

library(ggsignif)

# Plot with both effect predictions and raw datapoints
ggplot() +
  #geom_point(data = ppositive_graph_long, aes(x = time, y = alex_idcom_mean_score, group = participant_id,
  #color = ELA_group_w1_f), alpha = 0.7, position = position_jitter(width = 0.05, height = 0)) +
  #geom_line(data = ppositive_graph_long, aes(x = time, y = alex_idcom_mean_score, color = ELA_group_w1_f, group = participant_id), size = 1, alpha=0.5) +
  # Predicted effects
  geom_point(data = effects_poscom, aes(x = time, y = fit, group = poscom_c), size = 3) +
  geom_line(data = effects_poscom, aes(x = time, y = fit, group = poscom_c, linetype = poscom_c), size = 1) +
  scale_linetype_manual(
    name = "Simple Slopes",
    values = c("solid", "dotted")
  ) +
  xlab("Timepoint") + 
  scale_x_discrete(
  breaks = c('w1', 'w3'),
  labels = c("Baseline", "Followup"),
  expand = c(0, 0.3)
) + 
  ylab("Alexithymia") +
  labs(color = "crEA-Exposure Group") + 
  #scale_color_brewer(palette = "Spectral") +
  scale_color_manual(
    name = "crEA-Exposure Group",
    values = c("Comparison" = dark2_cols[1], "crEA-Exposed" = dark2_cols[2])
  ) +
  ggtitle("Positive Social Communication") +
  geom_signif(
  data = filter(poscom_graph_long, time == "w1"),
  aes(x = ELA_group_w1_f, y = alex_idcom_mean_score),
  comparisons = list(c("Comparison", "crEA-Exposed")),
  annotations = "*",
  y_position = 0.81,
  tip_length = 0.02
) +
  #annotate("text", x = "w1", y = 0.81, label = "*", size = 8, fontface = "bold") +
  theme_bw() + theme(panel.grid.major = element_blank(),
                         panel.grid.minor = element_blank(),
                         aspect.ratio = .6,
                         panel.border = element_rect(
                           color = "black",  
                           size = 0.5),
                         axis.title = element_text(size = 13),
                         axis.text = element_text(size = 11),
                        plot.title = element_text(hjust = 0.5, face="bold"),
                        legend.position = "bottom")
```

### Parenting Moderation Plots
```{r}
library(patchwork)

combined_plot <- ppos_plot + poscom_plot +
  plot_annotation(tag_levels = 'A') +
  plot_layout(ncol = 2 , guides = "collect") &
  theme(legend.position = "bottom")

ggsave("figures/parenting_plot.png", plot = combined_plot, width = 10, height = 6, dpi = 1000)
```


## 2d: MicroFIMS and Alexithymia
### Latent Growth Model: Baseline and Change over time differences
```{r}
model_pactive.cpositi <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ bo*1 + b_pc*pactive.cpositi + b_group*ELA_group_w1 + b_age*child_age_years_w1 + b_sex*female + b_race*primarycg_poc + b_format*format
  s ~ g1*1 + g_pc*pactive.cpositi  + g_group*ELA_group_w1 + g_age*child_age_years_w1 + g_sex*female + g_race*primarycg_poc + g_format*format

  # Covariance between growth factors (residual after prediction)
  i ~~ s
  '
  
fit_pactive.cpositi <- growth(model_pactive.cpositi,
                          data = sample,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_pactive.cpositi, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_pactive.cpositi, type = "std.all", ci=TRUE, level = .95)

# run model without ppostivie outliers
sample_nodyad_out <- sample %>%
  dplyr::filter(!(participant_id %in% conting_outlier_ids$participant_id))


fit_pactive.cpositi_noout <- growth(model_pactive.cpositi,
                          data = sample_nodyad_out,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")
summary(fit_pactive.cpositi_noout, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE) # no longer effect
```

# Exploratory Aim 3: Age Differences
## Age*crEA group on alexithymia: growth model 
```{r}
# mean center predictors
sample <- sample %>%
  mutate(
    group_c = as.vector(scale(ELA_group_w1, center=TRUE, scale=FALSE)),
    age_c = as.vector(scale(child_age_years_w1, center=TRUE, scale=FALSE)),
    # add interaction term to dataset
     group_x_age = group_c*age_c
  )

# add to lgm 
model_conditional_age <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ group_x_age + group_c + age_c + child_sex_w1
  s ~ group_x_age + group_c + age_c + child_sex_w1

  # Covariance between growth factors (residual after prediction)
  i ~~ s'

fit_conditional_age <- growth(model_conditional_age,
                          data = sample,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_conditional_age, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_conditional_age, type = "std.all", ci=TRUE, level = .95)
```

### Plot relationship between age and baseline alextihymia in crEA group
```{r}
# age and alexithymia at baseline in crEA group only
cor.test(sample_ela$alex_idcom_mean_score_w1, sample_ela$child_age_years_w1)

ggplot(sample_ela, aes(x = child_age_years_w1, y = alex_idcom_mean_score_w1)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Child Age (years)",
    y = "Baseline Alexithymia",
    title = "Age and Baseline Alexithymia in crEA Group"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

sample_ela %>% dplyr::count(!is.na(alex_idcom_mean_score_w1))

sample <- sample %>%
  mutate(
    ELA_group_w1_f = factor(
      ELA_group_w1,
      levels = c(0, 1),
      labels = c("Comparison", "crEA-Exposed")
    )
  )

# full sample with interaction
ggplot(sample, aes(
  x = child_age_years_w1,
  y = alex_idcom_mean_score_w1,
  color = ELA_group_w1_f
)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Age and Baseline Alexithymia in Each Group",
    x = "Child Age (years)",
    y = "Baseline Alexithymia",
    color = "ELA group"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

## Interaction of age * years in stable care on alexithymia
```{r}
# add interaction term to dataset
sample_ela <- sample_ela %>%
  mutate(
    years_sc_c = as.vector(scale(years_in_stable_care_new, center=TRUE, scale=FALSE)),
    age_c = as.vector(scale(child_age_years_w1, center=TRUE, scale=FALSE)),
    years_sc_x_age = age_c*years_sc_c
  )

# add to lavaan model
model_timeinstablecare_age <- '
  # Caregiving and covariates predicting growth factors
  alex_idcom_mean_score_w1 ~ years_sc_x_age + age_c + num_transitions_new + years_sc_c + child_sex_w1
'

fit_model_timeinstablecare_age <- sem(model_timeinstablecare_age,
                              data = sample_ela,
                              missing = "fiml",
                              estimator = "MLR",
                              cluster = "family_id")

summary(fit_model_timeinstablecare_age, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_model_timeinstablecare_age, type = "std.all", ci=TRUE, level = .95)
```

## Interaction of age * ss on alexithymia
```{r}
# mean center predictors
sample <- sample %>%
  mutate(
    ss_c = as.vector(scale(ss_score_w1, center=TRUE, scale=FALSE)),
    #already added: age_c = as.vector(scale(child_age_years_w1, center=TRUE, scale=FALSE)),
    # add interaction term to dataset
     ss_x_age = ss_c*age_c
  )

# add to lgm 
model_ss_conditional_age <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ ss_x_age + ss_c + ELA_group_w1 + age_c + child_sex_w1 + child_race_ethnicity
  s ~ ss_x_age + ss_c  + ELA_group_w1 + age_c + child_sex_w1 + child_race_ethnicity

  # Covariance between growth factors (residual after prediction)
  i ~~ s'

fit_ss_conditional_age <- growth(model_ss_conditional_age,
                          data = sample,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_ss_conditional_age, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_ss_conditional_age, type = "std.all", ci=TRUE, level = .95)
```

## Interaction of age * ppositive on alexithymia
```{r}
# add interaction term to dataset
sample <- sample %>%
  mutate(
    ppositive_c = as.vector(scale(ppositive, center=TRUE, scale=FALSE)),
    ppositive_x_age = age_c*ppositive_c
  )

# add to lgm
model_ppositive_age <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ bo*1 + b_ppa*ppositive_x_age + b_pp*ppositive_c + b_group*ELA_group_w1 + b_age*age_c + b_sex*female + b_race*primarycg_poc
  s ~ g1*1 + g_ppa*ppositive_x_age + g_pp*ppositive_c  + g_group*ELA_group_w1 + g_age*age_c + g_sex*female + g_race*primarycg_poc

  # Covariance between growth factors (residual after prediction)
  i ~~ s'
  
fit_ppositive_age <- growth(model_ppositive_age,
                          data = sample,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_ppositive_age, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_ppositive_age, type = "std.all", ci=TRUE, level = .95)
```

## Interaction of age * poscom on alexithymia
```{r}
# add centered vars to df
sample <- sample %>%
  mutate(
    poscom_c = as.vector(scale(poscom, center=TRUE, scale=FALSE)),
    poscom_x_age = age_c*poscom_c
  )

# add to lgm
model_poscom_age <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ bo*1 + b_pc*poscom_x_age + b_pp*poscom_c + b_group*ELA_group_w1 + b_age*age_c + b_sex*female + b_race*primarycg_poc
  s ~ g1*1 + g_pc*poscom_x_age + g_pp*poscom_c  + g_group*ELA_group_w1 + g_age*age_c + g_sex*female + g_race*primarycg_poc

  # Covariance between growth factors (residual after prediction)
  i ~~ s
  '
  
fit_poscom_age <- growth(model_poscom_age,
                          data = sample,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_poscom_age, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_poscom_age, type = "std.all", ci=TRUE, level = .95)
```

## Interaction of age * pactive.cpositiv on alexithymia
```{r}
# center vars
sample <- sample %>%
  mutate(
    pactive.cpositi_c = as.vector(scale(pactive.cpositi, center=TRUE, scale=FALSE)),
    pactive.cpositi_x_age = age_c*pactive.cpositi_c
  )

# add to lgm
model_pactive.cpositi_age <- '
  # Growth factors
  i =~ 1*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3
  s =~ 0*alex_idcom_mean_score_w1 + 1*alex_idcom_mean_score_w3

  # Residual variance constrained equal (required with 2 timepoints)
  alex_idcom_mean_score_w1 ~~ r*alex_idcom_mean_score_w1
  alex_idcom_mean_score_w3 ~~ r*alex_idcom_mean_score_w3

  # Caregiving and covariates predicting growth factors
  i ~ bo*1 + b_pca*pactive.cpositi_x_age + b_pc*pactive.cpositi_c + b_group*ELA_group_w1 + b_age*age_c + b_sex*female + b_race*primarycg_poc
  s ~ g1*1 + g_pca*pactive.cpositi_x_age + g_pc*pactive.cpositi_c  + g_group*ELA_group_w1 + g_age*age_c + g_sex*female + g_race*primarycg_poc

  # Covariance between growth factors (residual after prediction)
  i ~~ s
  '
  
fit_pactive.cpositi_age <- growth(model_pactive.cpositi_age,
                          data = sample,
                          missing = "fiml",
                          estimator = "MLR",
                          cluster = "family_id")

summary(fit_pactive.cpositi_age, fit.measures = TRUE, estimates = TRUE, rsquare = TRUE, standardized=TRUE)
standardizedSolution(fit_pactive.cpositi_age, type = "std.all", ci=TRUE, level = .95)
```